---
title: "Analysis of Health Care Access Dataset"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prepare Workspace, Load Data

```{r}
# Load data
load("data_cleaned.RData")

##library
library(ggplot2)
```


# Logistic Regression

## Fit Model to Full Dataset

```{r}
glm.fits <- glm(
y ~ .,
data = df , family = binomial 
)
summary(glm.fits)
```

## Interpret Model Summary
Which coefficients are reported as being statistically significant? 

Assuming $\alpha$ = 0.05

*note:* 
Variable’s p-values very small = variable associated with probability of binary response variable 
-> clear evidence of a real association between these variables and DiagPeriodL90D

The negative coeﬀicient for predictors suggests that the more likely the predictors occurs, then the person is less likely to have received treatment within the first 90 days since their diagnosis.

The positive coeﬀicient for predictors suggests that the more likely the predictors occurs, then the person is more likely to have received treatment within the first 90 days since their diagnosis.

```{r}

##subsetting data to only show coefficient values and p-value respectively 
co <- summary(glm.fits)$coef[, 1]
pVal <- summary(glm.fits)$coef[, 4]

##which variables' p-value is less than or equal to 0.05 (which is statistically significant?)
statSIG <- c()
coefSS <- c() 
for(i in 1:length(pVal)){
  if(pVal[i]<= 0.05){
    statSIG <- c(statSIG, pVal[i])
    coefSS <- c(coefSS, co[i])
  }
}

##if you have time: plot coefficient magnitudes
statSIG           ##clear evidence of a real association between these variables and DiagPeriodL90D
coefSS            ##the coefficient of these variables

#which statistically significant variables have negative coefficients (factors that correlate to lower likelihood of receiving treatment)?  
neg_coefSS <- coefSS[coefSS<=0]
neg_coefSS

#which statistically significant variables have positive coefficients (factors that correlate to higher likelihood of receiving treatment)?  
pos_coefSS <- coefSS[coefSS>=0]
pos_coefSS

##which statistically significant variable has the smallest coefficient? Ans: payer_typeMEDICARE ADVANTAGE
coefSS[which.min(coefSS)]

##which statistically significant variable has the biggest coefficient? Ans: bachelor degree 
coefSS[which.max(coefSS)]


##Which variable has the smallest p-value? (means anything??)
pVal[which.min(pVal)]           ##the payer type being commerical has the highest statistical significance 
co[which.min(pVal)]

```
# Coefficient Plots



```{r}
# Plot on log odds scale
signif_coefs <- as.data.frame(coef(summary(glm.fits)))
signif_coefs <- subset(signif_coefs, signif_coefs[, 4] < 0.05)
signif_coefs$Predictor <- rownames(signif_coefs)
ggplot((signif_coefs), aes(x = Predictor, y = Estimate)) +
  geom_bar(stat = "identity") +
  coord_flip()

```
```{r}
# Plot on odds scale
signif_coefs_odds <- signif_coefs
signif_coefs_odds$Estimate <- exp(signif_coefs$Estimate)
ggplot(signif_coefs_odds, aes(x = Predictor, y = Estimate)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_hline(yintercept = 1, color = "red")
```


## Generate Predicted Probabilities
```{r}
glm.probs <- predict(glm.fits, type = "response")

```


## Use Predicted Probabilities to Predict Response Values

Use a threshold of 0.5.

```{r}
glm.pred <- rep(0, 12873)
glm.pred[glm.probs > .5] = 1

attach(df)
##confusion matrix
cm <- table(glm.pred, y)

sensitivity <- (cm[2,2])/sum(cm[,2])
sensitivity         ##Logistic regression correctly predicted if someone get treatment  97.69% of the time.

specificity <- cm[1,1]/sum(cm[,1])
specificity         ##Logistic regression correctly predicted if someone does not get a treatment 3.97% of the time 

##has to match y level: training error rate  
training_error_rate <- mean(as.factor(glm.pred) == y) 

initial.pred <- data.frame(training_error_rate, specificity, sensitivity)
```


## Training Error Rate

Misclassification rate for the in-sample predictions (that is, the predictions made for the same data used to fit the model).

Training error rate is 62.5%

# Estimating the Test Error via 10-Fold Cross Validation
1. randomly split into 10 folds: sample(n) -> random permutation of 1, 2,..,n
  fold size = total observation / 10 
  a. have last fold = bigger 
2. for kth fold: training df = not kth fold  
  a. fit model on all data except kth fold 
  b. predict y for kth fold: error, specificity, sensitive (on kth fold) 

```{r}
storage <- as.data.frame(matrix(nrow = 10, ncol = 3))
n <- 12873
idx <- sample(n)
fold_size <- floor(n / 10)
for (i in 1:10) {
  # start and end positions of index vector for ith fold
  start_position <- (i - 1) * fold_size + 1
  end_position <- i * fold_size
  if (i == 10) {
    end_position <- n
  }
  # test indices (indices for ith fold)
  test_ind <- idx[start_position:end_position]
  
  # training indices (all indices not in ith fold)
  train_ind <- (1:n)[-test_ind]
  
  # fit model on all data except ith fold
  glm.fit_Train <- glm(y ~ ., data = df, subset = train_ind, family = binomial)
  
  # predict on ith fold
  pi_hat <- predict(glm.fit_Train, df[test_ind, ], type = "response")
  y_hat <- as.numeric(pi_hat > 0.5)
  
  # compute accuracy rates
  cmat <- table("Truth" = df[test_ind, "y"],
                "Pred" = y_hat)
  acc <- c("Overall" = sum(diag(cmat)) / sum(cmat), # Overall accuracy
           "Specificity" = cmat[1, 1] / sum(cmat[1, ]), # Specificity
           "Sensitivity" = cmat[2, 2] / sum(cmat[2, ])  # Sensitivity
  )
  storage[i, ] <- round(acc, 3)
  
}
names(storage) <- c("Overall Accuracy", "Specificity", "Sensitivity")
```


3. average error, specificity, sensitivity 
```{r}
cv_acc <- colMeans(storage)
names(cv_acc) <- c("Overall", "Specificity", "Sensitivity")
cv_acc
```


