---
title: "Cleaning of Health Care Access Dataset"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prepare Workspace, Load Data

```{r}
# Load packages
library(dplyr)  # for select()
library(ggcorrplot)
```

Data source: [WiDS (Women in Data Science) 2024 Datathon Challenge #1](https://www.kaggle.com/competitions/widsdatathon2024-challenge1/data)

From the source:

*"For this datathon challenge, we are using a real-world evidence dataset from Health Verity (HV), one of the largest healthcare data ecosystems in the US, as the main data source for the Datathon. In particular, the HV dataset that we use for this challenge contains health related information of patients who were diagnosed with metastatic triple negative breast cancers in the US. We also enriched the data set with the US Zip Codes Database which were built from the ground up using authoritative sources including the U.S. Postal Service™, U.S. Census Bureau, National Weather Service, American Community Survey, and the IRS, to obtain additional social economic information based on the locations of the patients. The dataset was then further enriched, also using zip code level, with toxicology data from NASA/Columbia University, to explore the relations between health outcomes and toxic air conditions."*

Note: we use only the training data set since it contains data for the response variable (for competition purposes, response values for the test data are not publicly available).

```{r}
# Load data
wids <- read.csv("data_raw.csv")
```


```{r}
# Initial data dimensions:
dim(wids)
```



```{r}
# Initial summary of variables:
str(wids)
```



# Key Variables of Interest

## Response

The response variable is `DiagPeriodL90D`, which is an indicator of whether the person received treatment within the first 90 days since their diagnosis.

For context, all subjects in the data set are patients who were diagnosed with metastatic triple negative breast cancer, an extremely aggressive type of cancer for which rapid treatment is a critical factor in the likelihood of survival. Due to this fact, such a patient failing to receive treatment within 90 days of their diagnosis is reasoned to be a suitable proxy for lacking access to health care.

The response variable is distributed as follows in our data:

```{r}
barplot(table(wids$DiagPeriodL90D) / sum(table(wids$DiagPeriodL90D)), main = "`DiagPeriodL90D` Distribution", xlab = "Diagnosis Within 90 Days", ylab ="Relative Frequency")

##convert to ggplot to save
rDist <- ggplot(data = as.data.frame(table(wids$DiagPeriodL90D) / sum(table(wids$DiagPeriodL90D))), aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity") +
  labs(title = "`DiagPeriodL90D` Distribution", x = "Diagnosis Within 90 Days", y = "Relative Frequency")

rDist

save(rDist, file = "response.rdata")
```

We see that the two levels have a reasonably balanced representation in our data (this is important to check, since issues can occur when one class contains very few observations).


## Patient Race

The predictor in which we are most interested is `patient_race`, which is distributed as follows in our data:

```{r}
barplot(table(wids$patient_race))
```

Around half of the observations have a blank value for patient race. The meaning of a blank value is unclear; however, since the coded values include a level for `Other`, a blank value most likely indicates that the subject declined to provide this information. 

Because the decision of whether or not to provide this information could be associated with the value of the response variable (e.g., the decision not to provide a racial category could be associated with one's race, which in turn could be associated with whether or not one receives treatment within 90 days), we should not remove the observations with blank patient race values. Instead, we leave them intact in the data set.



# Data Cleaning



## Removing Some Predictors

Predictors we will remove when performing standard logistic regression:

- Patient ID (not informative)
- Patient Gender (all subjects are female)
- Patient State
    - Categorical with 50+ levels
    - Not necessary to remove, but adding 49 indicator variables to the model is probably not ideal
    - Data does contain variables for larger geographic region (west/south/etc), so some geographic info is still present, although we lose some capacity to detect effects due to state-specific policy
    - We can include this variable if/when we do automatic variable selection with the Lasso; then the model will only include indicator variables for states with a significant effect 
- Patient ZIP code
    - Categorical with 739 levels
    - Can potentially include if/when using Lasso
- Breast cancer diagnosis code
    - Similar situation to patient state, categorical with 50 levels
    - Include if/when using Lasso for variable selection
- Breast cancer diagnosis desc
    - Duplicates information in Breast cancer diagnosis code
- Metastatic cancer diagnosis code
    - Similar situation to breast cancer diagnosis code (43 levels)
    - Include if/when using Lasso
- `metastatic_first_novel_treatment` and `metastatic_first_novel_treatment_type`
    - Only contains 24 non-blank values (not informative)

```{r}
table(wids$patient_gender)
```


```{r}
length(unique(wids$breast_cancer_diagnosis_code))
```


```{r}
length(unique(wids$metastatic_cancer_diagnosis_code))
```

```{r}
length(table(wids$patient_zip3))
```


```{r}
length(table(wids$metastatic_first_novel_treatment_type))
```

```{r}
table(wids$metastatic_first_novel_treatment)
```

```{r}
table(wids$metastatic_first_novel_treatment_type)
```

```{r}
df <- select(wids, !c(patient_id, patient_state, patient_zip3, patient_gender,
                      breast_cancer_diagnosis_code,
                      breast_cancer_diagnosis_desc,
                      metastatic_cancer_diagnosis_code,
                      metastatic_first_novel_treatment_type,
                      metastatic_first_novel_treatment))
```



## Numeric Types Stored As Character

Ozone, PM25 and NO2 contain numeric data but are stored as character vectors:

```{r}
df$Ozone <- as.numeric(df$Ozone)
df$PM25 <- as.numeric(df$PM25)
df$N02 <- as.numeric(df$N02)
```

## Missing Values

We check the number of missing values for each variable with more than five values missing:

```{r}
num_NAs <- sapply(df, function(x) sum(is.na(x)))
sort(num_NAs[num_NAs > 5])
```

Most observations are missing a value for `bmi`, so dropping all rows with one or more missing values would cause us to lose most of the data.

Meanwhile, the data is nearly complete in the rest of the variables. So if we instead drop `bmi`, then we need only remove a small share of data rows to eliminate the remaining missing values.

```{r}
# Drop BMI
df <- select(df, !bmi)

# Remove missing values
df <- na.omit(df)

# Resulting dimensions
dim(df)
```

## Convert Character Types to Factors

```{r}
# Convert character variables to factors
df[sapply(df, is.character)] <- lapply(df[sapply(df, is.character)], as.factor)

# Check final data types of predictors
str(df)
```


## Convert Response to Factor and Rename

```{r}
names(df)[names(df) == "DiagPeriodL90D"] <- "y"
df$y <- as.factor(df$y)
```



# Correlation Among Quantitative Predictors

```{r}
# Numeric variables only
df_num <- select(df, !c(patient_race, payer_type, Region, Division, y))
```

Subset rows (columns) with strong correlation for better visualization, due to the large number of predictors:

```{r}
# Matrix of pairwise correlations among predictors
xcor <- cor(as.matrix(df_num))

# IDs of variables having absolute correlation above 0.9 with at least one other 
ids <- apply(xcor, 2, function(x) { sum(abs(x) > 0.9) > 1 })

# Correlation heatmap for subset of variables with strong correlation
heatmap <- ggcorrplot(xcor[ids, ids], lab = TRUE, lab_size = 2)

save(heatmap,file="WiDS dataset.rdata")

```

For each pair of variables with over 0.9 correlation, remove one, repeating until no absolute correlations above 0.9 remain:

```{r}
df <- select(df, !c(
  housing_units,               # population
  male,                        # female
  married,                     # never_married
  income_individual_median,    # income_household_median
  income_household_six_figure, # income_household_median
  income_household_150_over,   # income_household_median
  home_value,                  # rent_median
  education_college_or_above,  # education_graduate
))
```



# Save Cleaned Data

```{r}
save(df, file = "data_cleaned.RData")
```


