---
title: "Analysis of Health Care Access Dataset"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prepare Workspace, Load Data

```{r}
library(glmnet)
library(reshape2)
library(dplyr)
library(ggplot2)
library(cowplot)
library(viridis)
library(grDevices)
```


```{r}
# Load data
load("data_cleaned.RData")
```



# Variable Selection via LASSO 

Perform variable selection by tuning the penalty parameter $\lambda$ to yield the smallest model that achieves Binomial Deviance (estimated using 10-fold cross validation) within one standard error of the minimum value over all considered values of $\lambda$.

```{r}
# Create design matrix for predictors (required for gv.glmnet() input format)
X_design <- model.matrix(y ~ ., df)[, -1]

# Fit logistic regression model with LASSO penalty for variable selection
# Automatically tries many different values of shrinkage parameter lambda
cv_lrlasso <- cv.glmnet(X_design, df$y, family = "binomial")
plot(cv_lrlasso)
```

This results in a model containing the following predictors:

```{r}
# cv_lrlasso$glmnet.fit$lambda contains the sequence of lambda values tried
# Get the index of the lambda value where deviance is 1 SE above minimum
# (corresponds to right-hand vertical dashed line in plot; left is optimum)
# (we accept slightly suboptimal deviance in exchange for fewer predictors)
idx_lambda <- with(cv_lrlasso, which.min(abs(glmnet.fit$lambda - lambda.1se)))

# Get the coefficient estimates corresponding to the selected lambda
beta_lambda <- cv_lrlasso$glmnet.fit$beta[, idx_lambda]       
lasso_coefs <- names(beta_lambda)[which(beta_lambda != 0)]    
lasso_coefs
```

Note: `education_less_highschool` and `education_bachelors` are zip-code level quantitative variables measuring percentage of population who belong to the named educational demographic.

We fit a logistic regression model with the selected predictors to the full sample of observations, and we use 10-fold cross validation to estimate the overall and class-specific accuracy rates.

```{r}
# Create input data corresponding to variables selected by LASSO
dfsub <- df
dfsub$patient_raceWhite <- df$patient_race == "White"
dfsub$payer_typeCOMMERCIAL <- df$payer_type == "COMMERCIAL"
dfsub$RegionWest <- df$Region == "West"
dfsub <- select(dfsub, all_of(lasso_coefs))
dfsub$y <- df$y# Create design matrix for predictors (required for gv.glmnet() input format)

# Fit logistic regression model using variables selected by LASSO
fit_lrlasso <- glm(y ~ ., data = dfsub, family = binomial)

# Generate predictions using fitted model
pi_hat <- predict.glm(fit_lrlasso, dfsub, type = "response")
y_hat <- as.numeric(pi_hat > 0.5)

# Create confusion matrix
cmat <- table("Truth" = df$y, "Pred" = y_hat)
cmat
```



```{r}
# Accuracy rates
acc <- c("Overall" = sum(diag(cmat)) / sum(cmat), # Overall accuracy
         "Specificity" = cmat[1, 1] / sum(cmat[1, ]), # Specificity
         "Sensitivity" = cmat[2, 2] / sum(cmat[2, ])  # Sensitivity
)
acc <- round(acc, 3)
acc
```



```{r}
# Coefficient estimates
summary(fit_lrlasso)
```

```{r}
# Coefficient estimates on odds-ratio scale instead of log odds ratio
exp(coef(summary(fit_lrlasso)))
```

E.g., a white person has ~1.14 times the odds of treatment compared to a non-white person.



# Comparison to Significant Predictors in Full Model

```{r}
# fit Logistic Regression with all predictors
fit_lrfull <- glm(y ~ ., data = df, family = binomial)

# Generate predictions using fitted model
pi_hat <- predict.glm(fit_lrfull, df, type = "response")
y_hat <- as.numeric(pi_hat > 0.5)

# Create confusion matrix
cmat_full <- table("Truth" = df$y, "Pred" = y_hat)
cmat_full

# Accuracy rates
acc_full <- c("Overall" = sum(diag(cmat_full)) / sum(cmat_full), # Overall accuracy
         "Specificity" = cmat_full[1, 1] / sum(cmat_full[1, ]), # Specificity
         "Sensitivity" = cmat_full[2, 2] / sum(cmat_full[2, ])  # Sensitivity
)
acc_full <- round(acc_full, 3)
acc_full
```


```{r}
# Significant predictors
signif_coefs <- coef(summary(fit_lrfull))
signif_coefs <- subset(signif_coefs, signif_coefs[, 4] < 0.05)
signif_coefs
```

```{r}
# Coefficient estimates on odds-ratio scale instead of log odds ratio
exp(signif_coefs[, 1])
```




# Coefficient Plots



```{r}
# Plot on log odds scale
signif_coefs <- as.data.frame(coef(summary(fit_lrfull)))
signif_coefs <- subset(signif_coefs, signif_coefs[, 4] < 0.05)
signif_coefs$Predictor <- rownames(signif_coefs)
ggplot((signif_coefs), aes(x = Predictor, y = Estimate)) +
  geom_bar(stat = "identity") +
  coord_flip()

```
```{r}
# Plot on odds scale
signif_coefs_odds <- signif_coefs
signif_coefs_odds$Estimate <- exp(signif_coefs$Estimate)
ggplot(signif_coefs_odds, aes(x = Predictor, y = Estimate)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_hline(yintercept = 1, color = "red")
```



```{r}
# compare coefficient estimates
make_coefplot(fit_lrfull, fit_lrlasso)
```


# Cross-Validation (Without LASSO)


```{r}
storage <- matrix(nrow = 10, ncol = 3)
n <- 12873
idx <- sample(n)
fold_size <- floor(n / 10)
for (i in 1:10) {
  # start and end positions of index vector for ith fold
  start_position <- (i - 1) * fold_size + 1
  end_position <- i * fold_size
  if (i == 10) {
    end_position <- n
  }
  # test indices (indices for ith fold)
  test_ind <- idx[start_position:end_position]
  
  # training indices (all indices not in ith fold)
  train_ind <- (1:n)[-test_ind]
  
  # fit model on all data except ith fold
  glm.fit_Train <- glm(y ~ ., data = df, subset = train_ind, family = binomial)
  
  # predict on ith fold
  pi_hat <- predict(glm.fit_Train, df[test_ind, ], type = "response")
  y_hat <- as.numeric(pi_hat > 0.5)
  
  # compute accuracy rates
  cmat <- table("Truth" = df[test_ind, "y"],
                "Pred" = y_hat)
  acc <- c("Overall" = sum(diag(cmat)) / sum(cmat), # Overall accuracy
           "Specificity" = cmat[1, 1] / sum(cmat[1, ]), # Specificity
           "Sensitivity" = cmat[2, 2] / sum(cmat[2, ])  # Sensitivity
  )
  storage[i, ] <- round(acc, 3)
  
}
```

```{r}
cv_acc <- colMeans(storage)
names(cv_acc) <- c("Overall", "Specificity", "Sensitivity")
cv_acc
```


```{r}
table(df$y)/sum(table(df$y))
```

